Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-26 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update draft (intro)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2899 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5026cbd44fb6b5144594f5f292b335b0a2eac778

parameter constant. Additionally, they introduce a combined hetereoscedastic version 

@@

complexity encountered in optimizing the log-likelihood function which is not globally concave. 
Therefore, highly informative starting values are crucial for the model to converge at all 

@@

present an alternative approach to regression modeling of circular data by employing a GLM in a 
Bayesian framework, however, only the mean direction is regressed on covariates.

@@

Generally these provide a powerful toolbox for modeling particularly smooth effects,
however, in the case of circular data a possibly highly non-linear link function
can induce a more complex behavior of the modeled parameters as the location
parameter shifts closely towards the ``origin point'' $0$, also depending on the 
direction of the shift.
Moreover, not all types of effects may be modeled adequately. Especially abrupt changes 

@@

%% FIXME: clear explanation of difficulties close to 0

In order to avoid these difficulties in probabilistic regression modeling of circular data 
(and also the requirement of initial values) we propose an alternative and very flexible 
approach by applying the von Mises distribution within the methodology of distributional 
trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests 
(for short: \emph{circular trees and forests}) 
do not require any model assumptions as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while 
circular forests allow for modeling smooth changes as well. Thus, a high level of 
flexibility is provided while a full distributional model is specified in each terminal 
node offering a wide range of inference. Moreover, due to the parameters being estimated 
as constants for each subgroup difficulties connected to ``circular shifts'' close to $0$ 
are avoided.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This novel approach to circular regression trees and forests complements the literature 
on tree-based circular modeling.
%In particular, 
\cite{Lund:2002} have already introduced the concept of circular regression 
trees where binary splits are made based on a circular distance which is evaluated between 
the observed response values and the conditional mean direction of the circular response variable
and in that way serves as a measure of homogeneity within subgroups. 
%For a circular response $\theta$, covariate(s) $z$ and the conditional mean direction $\mu(z)$
%the employed circular distance is defined by $1-\cos(\theta - \mu(z)$.
%In each terminal node a predicted values is provided by the sample mean direction of the corresponding 
%subset of observations.
While this approach and also further extensions \citealp{Larraondo+Inza+Lozano:2018} allow not only for 
linear but also for a circular covariate, the tree splits are performed based on effects in the expected 
value only, i.e., only the location parameter $\mu$ is regressed on covariates.
%Furthermore, even though an estimate of dispersion in the form of total node heterogeneity is provided 
%along with the predicted mean value in each terminal node
Hence, the here proposed circular regression trees and forests can be seen as a further step in the development
of tree-based circular models by incorporating the framework of distributional modeling and in that way
considering splits in all distribution parameters and providing a fully specified distribution fitted 
in each terminal node.

>>>>

parameter constant. Additionally, they developed a combined hetereoscedastic version 

@@

complexity encountered in optimizing the corresponding log-likelihood function which is not globally concave. 
Therefore, highly informative starting values are crucial for these circular GLMs to converge at all 

@@

present a Bayesian alternative of a homoscedastic GLM for circular data.

@@

Generally these provide a powerful toolbox for modeling particularly smooth effects.
However, in the case of modeling circular data 
addititve effects conditional on linear covariates are not intuitive
due to a possibly strongly non-linear rotation on the unit circle, whereby the same transformation
can be obtained in either positive or negative direction on the circle.
%a possibly highly non-linear link function
%can induce a more complex behavior of the modeled parameters as the location
%parameter shifts closely towards the ``origin point'' $0$, also depending on the 
%direction of the shift.
Moreover, abrupt changes 

@@

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
%In order to avoid these difficulties in regression modeling of circular data 
%(and also the requirement of initial values) 
%we propose an alternative and very flexible 
%approach by applying the von Mises distribution within the methodology of distributional 
%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests 
do not require a pre-specified link function or any other model assumptions as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while 
forests allow for modeling smooth changes as well. Thus, a high level of 
flexibility is provided while a full distributional model is specified in each terminal 
node offering a wide range of inference. Moreover, as covariates are employed for splitting only and not
considered for the parameter estimation within the resulting subgroups, their effects are easier to be interpreted for circular responses.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This novel approach to circular regression trees and forests complements the literature 
on tree-based circular modeling.
%In particular, 
\cite{Lund:2002} has already introduced the concept of circular regression 
trees where binary splits are made based on an angular distance measure employed to evalute node homogeneity.
%which is evaluated between 
%the observed response values and the conditional mean direction of the circular response variable
%and in that way serves as a measure of homogeneity within subgroups. 
%%For a circular response $\theta$, covariate(s) $z$ and the conditional mean direction $\mu(z)$
%%the employed circular distance is defined by $1-\cos(\theta - \mu(z)$.
%%In each terminal node a predicted values is provided by the sample mean direction of the corresponding 
%%subset of observations.
While this approach and also further extensions \citep{Larraondo+Inza+Lozano:2018} allow not only for 
linear but also for a circular covariate, the tree only considers effects in the mean direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
%Furthermore, even though an estimate of dispersion in the form of total node heterogeneity is provided 
%along with the predicted mean value in each terminal node
By incorporating the framework of distributional modeling, we extend the idea of circular trees allowing splits in all distribution parameters and providing a fully specified circular response distribution for each terminal node.

%Hence, the here proposed circular regression trees and forests can be seen as a further step in the development
%of tree-based circular models by incorporating the framework of distributional modeling and in that way
%considering splits in all distribution parameters and providing a fully specified distribution fitted 
%in each terminal node.
